---
title: "In-class Exercise 3: Analytical Mapping"
date: "26 January 2023"
date-modified: "`r Sys.Date()`"
execute: 
  echo: true
  eval: true
  message: false
  warning: false
editor: visual
---

# Loading packages and data

```{r}
pacman::p_load(tmap, tidyverse, sf)
```

## Import data

```{r}
NGA_wp <- read_rds("data/rds/NGA_wp.rds")
```

# Basic Choropleth Mapping 

Distribution of non-functional water points by LGA

```{r}
p1 <- tm_shape(NGA_wp) +
  tm_fill("wp_functional",
          n = 10,
          style = "equal",
          palette = "Blues") +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Distribution of functional water point by LGAs",
            legend.outside = FALSE)
```

```{r}
p2 <- tm_shape(NGA_wp) +
  tm_fill("total_wp",
          n = 10,
          style = "equal",
          palette = "Blues") +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Distribution of total  water point by LGAs",
            legend.outside = FALSE)
```

Explanation: tm_shape(spatial data)

```{r}
tmap_arrange(p2, p1, nrow = 1)
```

# Choropleth Map for Rates

Compare the proportion of functional water points to the proportion of non-functional water points.

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional/total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional/total_wp)
```

## Plotting map of rate

```{r}
tm_shape(NGA_wp) +
  tm_fill("pct_functional",
          n = 10,
          style = "equal",
          palette = "Blues",
          legend.hist = TRUE) +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Rate map of functional water point by LGAs",
            legend.outside = TRUE)
```

# Extreme Value Maps 

Identify outliers by highlighting extreme values at the lower and upper end of the scale.

## Percentile map

### Data Preparation

(1) Exclude records with NA:

```{r}
NGA_wp <- NGA_wp %>%
  drop_na()
```

(2) Create customers classification and extract values

```{r}
percent <- c(0,.01,.1,.5,.9,.99,1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[,1], percent)
```

### Creating functions

#### get.var function

R function to extract a variable as a vector out of an sf dataframe.

-   arguments:

    -   vname (as character, in quotes)
    -   df: name of sf data frame

-   returns

    -   v: vector with values (without a column name)

    ```{r}
    get.var <- function(vname,df) {
      v <- df[vname] %>% 
        st_set_geometry(NULL)
      v <- unname(v[,1])
      return(v)
    }
    ```

#### Percentile mapping function

```{r}
percentmap <- function(vnam, df, legtitle=NA, mtitle="Percentile Map"){
  percent <- c(0,.01,.1,.5,.9,.99,1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)
  tm_shape(df) +
  tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,
             title=legtitle,
             breaks=bperc,
             palette="Blues",
          labels=c("< 1%", "1% - 10%", "10% - 50%", "50% - 90%", "90% - 99%", "> 99%"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("right","bottom"))
}
```

**Test the percentile mapping function**

```{r}
percentmap("wp_nonfunctional", NGA_wp)
```

## Box map

```{r}
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()
```

### boxbreaks function

-   arguments:
    -   v: vector with observations
    -   mult: multiplier for IQR (default 1.5)
-   returns:
    -   bb: vector with 7 break points compute quartile and fences

```{r}
boxbreaks <- function(v,mult=1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode="numeric",length=7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) { # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

### get.var function

-   arguments:
    -   vname: variable name (as character, in quotes)
    -   df: name of sf data frame
-   returns:
    -   v: vector with values (without a column name)

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

**Test the functions**

```{r}
var <- get.var("wp_nonfunctional", NGA_wp) 
boxbreaks(var)

```

## Boxmap function

-   arguments:
    -   vname: variable name (as character, in quotes)
    -   df: simple features polygon layer
    -   legtitle: legend title
    -   mtitle: map title
    -   mult: multiplier for IQR
-   returns:
    -   a tmap-element (plots a map)

```{r}
boxmap <- function(vnam, df, 
                   legtitle=NA,
                   mtitle="Box Map",
                   mult=1.5){
  var <- get.var(vnam,df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,title=legtitle,
             breaks=bb,
             palette="Blues",
          labels = c("lower outlier", 
                     "< 25%", 
                     "25% - 50%", 
                     "50% - 75%",
                     "> 75%", 
                     "upper outlier"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("left",
                               "top"))
}
```

```{r}
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```

### Recode zero

Recode LGAs with zero total water point into NA:

```{r}
NGA_wp <- NGA_wp %>%
  mutate(wp_functional = na_if(
    total_wp, total_wp < 0))
```
