---
title: "In-class Ex 5: Advanced Spatial Point Pattern Analysis"
date-modified: "`r Sys.Date()`"
date: "6 February 2023"
execute: 
  message: false
  warning: false
editor: visual
---

#### Installing packages

sfdep: spatial and spatial time cubes, colocation purposes

```{r}
pacman::p_load(tidyverse, sf, tmap, sfdep)
```

#### Importing Data
Transform to Taiwan's projection System (3829)

```{r}
studyArea <- st_read(dsn = "data",
                    layer = "study_area") %>%
  st_transform(crs = 3829)
```

```{r}
stores <- st_read(dsn = "data", 
                  layer = "stores") %>%
  st_transform(crs = 3829)
```
## Visualising the sf layers 
```{r}
tmap_mode('view')
tm_shape(studyArea) + 
  tm_polygons() + 
  tm_shape(stores) + 
  tm_dots(col = "Name", 
          size = 0.01, 
          border.col = "black", 
          border.lwd = 0.5) + 
  tm_view(set.zoom.limits = c(12, 16))
```

## Local Coloation Quotients (LCLQ)
Calculating nearest neighbour list: 
```{r}
nb <- include_self(
  st_knn(st_geometry(stores), 6))
# 6 means to search for the 6 nearest neighbours
# - Adaptive method
# - Total 7 points (including self) so that you can get a good split. Avoid even splits e.g. 3, 3
```
```{r}
wt <- st_kernel_weights(nb, 
                        stores, 
                        "gaussian",
                        adaptive = TRUE
                        )
```

```{r}
FamilyMart <- stores %>%
  filter(Name == "Family Mart") # Select stores where name = 'Family Mart', in dataframe form 
A <- FamilyMart$Name #vector, target
```

```{r}
SevenEleven <- stores %>%
  filter(Name == "7-Eleven")
B <- SevenEleven$Name
```

Run 50 simulations: 
```{r}
LCLQ <- local_colocation(A, B, nb, wt, 49)
```

Combine stores and local location table.
-   Not doing left join / right join because these two layers have no unique identifiers. 
-   Don't sort data by code
-   Left hand side should be the geomerty, 'stores'
```{r}
LCLQ_stores <- cbind(stores, LCLQ) 
```

```{r}
tmap_mode('view')
tm_shape(studyArea) + 
tm_polygons() + 
  tm_shape(LCLQ_stores) + 
  tm_dots(col = "X7.Eleven", 
          size = 0.01, 
          border.col = "black", 
          border.lwd = 0.5) + 
  tm_view(set.zoom.limits = c(12, 16))
tmap_mode('plot')
```






